---
title: "A network-based approach to characterize heterogeneous bioinformatics training needs: Analysis of the SIB training survey"
bibliography: references.bib
author: Valeria Di Cola, Monique Zahn, Diana Marek, Gregoire Rossier, Patricia M. Palagi, Geert van Geest​
format:
  html:
    code-fold: true
    theme: cosmo
    embed-resources: true
    number-sections: true
  docx:
    number-sections: true
    reference-doc: custom-reference-doc.docx
    fig-format: png
    fig-dpi: 100
    fig-width: 4
execute:
  echo: false
  warning: false
  message: false
abstract: | 
  **Background:** Bioinformatics training audiences are heterogeneous, and demographic groupings alone often mask meaningful differences in training priorities. To address this, we analyzed the SIB training needs survey using a network-based strategy that identifies respondent subgroups from shared topic preferences, alongside standard analyses of barriers and learning formats.

  **Methods:** We analyzed 260 responses collected over 6 weeks (March-April 2026), including Likert-type ratings for training topics, participation barriers, and preferred course formats, plus demographic and course-attendance variables. Respondents were represented as nodes in a correlation network built from course-topic ratings and Louvain community detection was applied to identify clusters. Cluster profiles were then characterized using topic priorities, barriers, format preferences, and demographics.

  **Results:** Most responses came from Switzerland (67.3%) and academia (82.7%). Across all respondents, strongest demand centered on AI/LLM-related skills, FAIR and reproducible data practices, and computational methods. Distance and course fees emerged as the most relevant barriers, while blended, interactive, and multi-day formats were generally preferred; AI-assisted learning formats were less favored. Network analysis identified four interpretable learner profiles: AI and data, infectious disease genomics, functional genomics, and Biochemistry, imaging & AI.

  **Conclusions:** Network-based clustering revealed actionable training-need heterogeneity that is not captured by demographic stratification alone. Integrating these learner profiles into curriculum planning can improve course targeting, format selection, and barrier mitigation, supporting more effective and responsive bioinformatics training programs.
---

```{r}
#| label: setup
library(dplyr)
library(tidyr)
library(stringr)
library(forcats)
library(ggplot2)
library(jsonlite)
library(readxl)
library(purrr)
library(scales)
library(patchwork)
library(svglite)
library(glue)

source("functions.R")

excel_path <- "data/SIB Training survey.xlsx"
metadata_path <- "question_metadata.json"

survey_results <- readxl::read_excel(excel_path)

# remove leading/trailing whitespace from column names
colnames(survey_results) <- str_trim(colnames(survey_results))

survey_results$region <- ifelse(
  survey_results$`In which country do you work?` == "Switzerland",
  "Switzerland",
  "Other"
)

raw_meta <- jsonlite::read_json(metadata_path, simplifyVector = FALSE)$questions
meta_df <- purrr::map_dfr(raw_meta, \(q) {
  tibble::tibble(
    main_question = q$main_question_stem,
    question_meta_type = q$type,
    scale = list(unlist(q$scale))
  )
})

org_type_values <- c(
  "Academia/ Research institution",
  "Healthcare",
  "Industry",
  "Non-profit organisation"
)

# Career stages used in Likert breakdowns; anything else (including Bachelor
# and Master's students) is recoded to "Other" in the single mutate below.
career_likert_levels <- c(
  "PhD candidate",
  "Postdoctoral researcher",
  "Senior scientist/ Principal investigator",
  "Research assistant/ Technician",
  "Other"
)

survey_results <- survey_results |>
  dplyr::mutate(
    `Which career stage are you currently in?` = dplyr::if_else(
      `Which career stage are you currently in?` %in% career_likert_levels,
      `Which career stage are you currently in?`,
      "Other"
    ),
    `How would you describe the organisation in which you currently work?` = dplyr::if_else(
      `How would you describe the organisation in which you currently work?` %in%
        org_type_values,
      `How would you describe the organisation in which you currently work?`,
      "Other"
    )
  )

question_dictionary <- build_question_dictionary(survey_results, meta_df)

survey_long <- survey_results |>
  dplyr::mutate(respondent_id = dplyr::row_number()) |>
  dplyr::mutate(dplyr::across(-respondent_id, as.character)) |>
  tidyr::pivot_longer(
    cols = -respondent_id,
    names_to = "raw_column",
    values_to = "answer"
  ) |>
  dplyr::left_join(question_dictionary, by = "raw_column") |>
  dplyr::mutate(answer = dplyr::na_if(answer, "")) |>
  dplyr::mutate(
    answer = ifelse(answer == "Highly needed", "Very needed", answer)
  )

swiss_participants <- survey_long |>
  dplyr::filter(
    main_question == "In which country do you work?",
    answer == "Switzerland"
  ) |>
  dplyr::pull(respondent_id)

all_topic_questions <- c(
  "Data management and knowledge representation: specify your training needs You can skip topics that are not relevant",
  "Computational methods and AI: specify your training needs You can skip topics that are not relevant",
  "Omics analysis: specify your training needs You can skip topics that are not relevant",
  "Biomedicine and pathogens: specify your training needs You can skip topics that are not relevant",
  "Biodiversity and Ecology: specify your training needs You can skip topics that are not relevant"
)

sib_course_values <- c(
  "I have never attended a SIB course",
  "More than a year ago",
  "Within the last year"
)
```

## Introduction

The rapid and continuous evolution of high-throughput experimental technologies, large-scale data infrastructures, and data management has made bioinformatics training a necessity for life science researchers [@via_ten_2011]. National and international training providers, such as the SIB Swiss Institute of Bioinformatics, aim to provide high-quality, impactful courses to a heterogeneous audience. This audience ranges from academic PhD candidates requiring upskilling in specific analytical competences to corporate senior scientists or clinical personnel seeking broad overviews of emerging computational resources.

Training needs can be evaluated based on three dimensions: course topics, format, and logistics. The course topics address educational needs, while format involves alignment to preferences regarding learning formats (e.g., intensive block courses, problem-based learning, self-paced e-learning). Logistics involves mitigating systemic barriers to attendance, including geographical distance and financial restrictions. Previous training needs surveys have analyzed these dimensions independently or slice responses using classical demographic categories like career stage or geographic region [@attwood_global_2019; @drew_revisiting_2023]. While informative, such an approach assumes that individuals within a demographic sub-group are homogeneous. However, a post-doctoral researcher in structural biology may share more training needs with an industry mass-spectrometry specialist than with a post-doctoral researcher in plant ecology. 

Clustering of survey responses can be highly valuable for understanding respondent heterogeneity. This approach was already introduced in the eighties for market segmentation [@punj_cluster_1983], and is still frequently used. For example, for understanding heterogeneity in citizenship norms [@hooghe_comparative_2016], health literacy [@cheng_measuring_2024; @hyde_electronic_2019], bullying [@chung_are_2020], or company training needs [@wenger_analysing_2024]. For our case of bioinformatics training needs, identifying clusters of survey respondents with similar needs can provide a better understanding of response heterogeneity. Applying such a data-driven approach can help training providers design curricula that are tailored to the specific needs of these sub-groups, rather than relying on broad demographic categories.

In recent years, the fast growth of Artificial Intelligence (AI) and Large Language Models (LLMs) has transformed both the bioinformatics landscape and the nature of computational training itself [@wen_bin_goh_rethinking_2026; @ruan_large_2026]. AI is no longer just an advanced topic for specialized computer scientists; tools like GitHub Copilot and ChatGPT have shifted how researchers write code, automate workflows and retrieve data [@jiang_survey_2026]. LLM and generative AI are heavily used by scientists [@daepp_ai_2026], and they are frequently used for education [@bewersdorff_taking_2025], both in the classroom and for self-directed learning [@esiyok_acceptance_2025; @wu_unlocking_2024]. To understand the impact of AI on training needs, it is therefore important to identify what kind of AI-related topics are required by scientists. This can help training providers to design courses that address the specific needs, and to ensure that AI-related topics are integrated into the curriculum in a way that is relevant and useful for all learners.  

This paper describes a data-driven framework for mapping training needs along the three dimensions of course topics, format, and logistics using graph-based clustering. By converting individual responses into a correlation graph, we define profiles of learners who share common training needs, regardless of their institutional or regional backgrounds. We detail the underlying data preprocessing pipeline, a conceptual overview of graph construction, and subsequent characterization of the isolated learner profiles with a strong emphasis on how AI needs shape these profiles.

## Methods 

### Survey design and destribution

The SIB training needs survey collected information using a multi-tiered survey composed of main questions (acting as structural stems) and specific sub-questions representing individual items or technical sub-topics. All survey questions can be found on Zenodo **link**. In summary, the questionnaire evaluated course topics, barriers and formats using 5-point Likert-type scales:

* Training needs: Respondents evaluated their structural training requirement for detailed sub-topics nested within five global computational domains: *Omics analysis*, *Computational methods and AI*, *Data management and knowledge representation*, *Biomedicine and pathogens*, and *Biodiversity and Ecology*. The evaluation metric utilized an ordinal scale ranging from low necessity to urgent application needs (*Not needed* to *Absolutely essential*).
* Perceived participation barriers: Structural and logistical impediments were quantified using an ordinal scale anchored by *No barrier* and *Critical barrier*. Evaluated factors included course fees, travel distance, timing constraints, and prerequisite alignment.
* Preferred learning formats: Pedagogical format delivery options (e.g., webinars, bring your own data, problem-based learning, multi-day intensive courses, self-paced e-learning, and AI-assisted learning) were rated via a preference scale ranging from *Strongly dislike* to *Strongly like*.

Demographic covariates were also collected to characterize the respondent pool: career stage, geographic demographics (working country and Swiss canton), organization type (Academia, Healthcare, Industry, Non-profit), and historical interaction (time elapsed since attending their last SIB training course).

The survey was distributed through the SIB training mailing list, the SIB network and social media. The survey was open for responses for a period of 6 weeks, between March 17th and April 30th 2026, and participation was voluntary and anonymous.

### Data preprocessing and missing value handling

For the questions on training needs, respondents were explicitly permitted to skip technical topics that were not relevant to their work. Consequently, non-responses within the training needs questions were treated as meaningful omissions, i.e. 'not relevant'. To preserve the complete respondent pool and avoid the biases, missing values were therefore imputed with the minimum ordinal score of 1, corresponding to a *"Not needed"* response. All data analysis steps were implemented in R ([@rcoreteam_r_2026]; version 4.3.1) using the packages `dplyr` [@wickham_dplyr_2026] for data manipulation and `ggplot2` [@wickham_ggplot2_2016] for visualization.

### Graph-based network and community detection

To define clusters of participants without relying on pre-existing demographic labels, we modeled the survey respondents using network analysis (@fig-cluster-workflow). In this network, each individual respondent is represented as a single node. For every possible pair of respondents, we calculated the Pearson correlation coefficient based on their responses across all technical training sub-topics. This score reflects the similarity in their training needs. These similarity scores form the connections (edges) between nodes, where the weight of the connection represents the strength of the positive correlation between those two responses. To eliminate weak background noise and prevent the network from becoming too dense, we filtered based on a threshold of the Pearson correlation coefficient. We calculated the 80th percentile of all positive correlations (corresponding to a correlation coefficient of 0.32) across the dataset and pruned any connection falling below this threshold. Unsupervised clustering of the filtered respondent network was performed using the Louvain community detection algorithm [@blondel_fast_2008] using a resolution of 1. The network analysis and community detection was conducted using the `igraph` package in R [@csardi_igraph_2026]. 

```{mermaid}
%%| label: fig-cluster-workflow
%%| fig-cap: "Graph-based clustering workflow for identifying learner archetypes. Respondents are modeled as nodes in a network, with edges representing pairwise correlation of training needs. Louvain community detection is applied to identify clusters of respondents with similar training needs."

graph TD
    %% Define Styles
    classDef default fill:#f9f9f9,stroke:#333,stroke-width:1px;
    classDef process fill:#e1f5fe,stroke:#0288d1,stroke-width:1px,font-style:italic;
    classDef cluster fill:#fff,stroke:#333,stroke-width:2px;

    %% Nodes
    A[SURVEY RESPONDENT POOL]
    B(Pearson Pairwise Correlation)
    C[PRUNED RESPONDENT NETWORK <br> r > 0.32]
    D(Louvain Community Detection)
    
    E[Cluster 1 <br> AI & Data]
    F[Cluster 2 <br> Infectious diseases genomics]
    G[Cluster 3 <br> Functional genomics]
    H[Cluster 4 <br> Biochemistry, imaging and AI]

    %% Class Assignments
    class B,D process;
    class E,F,G,H cluster;

    %% Connections
    A --> B
    B --> C
    C --> D
    D --> E
    D --> F
    D --> G
    D --> H

```

## Results

### Demographics and overall responses

In total 260 responses were gathered, of which 175 (67.3%) were from Switzerland (@fig-demographics A). Most respondents from Switzerland were spread across the cantons with major universities (@fig-demographics B). Most respondents were working in academia (n = 215, 82.7%; @fig-demographics C). The career stage was distributed across academic professions (@fig-demographics D), with a majority of PhD candidates (n = 77, 26.6%) and senior scientists/PIs (n = 67, 25.7%).

```{r}
#| label: fig-demographics
#| fig-cap: "Respondent demographics. (A) Country of work. (B) Canton (if working in Switzerland). (C) Organisation type. (D) Career stage."

p_a <- survey_results |>
  rename(country = `In which country do you work?`) |>
  filter(!is.na(country)) |>
  count(country) |>
  filter(n >= 3) |>
  mutate(country = fct_reorder(country, n)) |>
  ggplot(aes(x = n, y = country)) +
  geom_col() +
  geom_text(aes(label = n), hjust = -0.2, size = 3.5) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.2))) +
  labs(x = "Count", y = NULL) +
  theme_minimal()

p_b <- survey_results |>
  rename(canton = `In which canton do you work?`) |>
  filter(!is.na(canton)) |>
  count(canton) |>
  mutate(canton = fct_reorder(canton, n)) |>
  ggplot(aes(x = n, y = canton)) +
  geom_col() +
  geom_text(aes(label = n), hjust = -0.2, size = 3.5) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(x = "Count", y = NULL) +
  theme_minimal()

p_c <- survey_results |>
  rename(
    org = `How would you describe the organisation in which you currently work?`
  ) |>
  filter(!is.na(org)) |>
  count(org) |>
  mutate(org = fct_reorder(org, n)) |>
  ggplot(aes(x = n, y = org)) +
  geom_col() +
  geom_text(aes(label = n), hjust = -0.2, size = 3.5) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.2))) +
  labs(x = "Count", y = NULL) +
  theme_minimal() +
  scale_y_discrete(labels = label_wrap(25))

p_d <- survey_results |>
  rename(career_stage = `Which career stage are you currently in?`) |>
  filter(!is.na(career_stage)) |>
  count(career_stage) |>
  mutate(career_stage = fct_reorder(career_stage, n)) |>
  ggplot(aes(x = n, y = career_stage)) +
  geom_col() +
  geom_text(aes(label = n), hjust = -0.2, size = 3.5) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(x = "Count", y = NULL) +
  theme_minimal() +
  scale_y_discrete(labels = label_wrap(25))

(p_a + p_b) /
  (p_c + p_d) +
  plot_layout(guides = "collect") +
  plot_annotation(tag_levels = "A") &
  theme(legend.position = "bottom")
```

For a wide range of pre-defined course topics, the respondents were asked to indicate their training needs on a 5-point Likert scale. The overall distribution of responses is shown in @fig-likert-preferences, which indicates that the majority of respondents expressed a need for training in AI-related topics, FAIR data and computational methods, as well as in integrating imaging with omics data. In contrast, topics related to e.g. transposable elements analysis and 16S sequencing data analysis were generally considered as less needed. 

```{r}
#| label: fig-likert-preferences
#| fig-cap: "Overall responses of training needs, shown in percentages and ordered by need. The topics are ordered by the proportion of respondents who indicated that they needed the topic (i.e. answered \"Very needed\" or \"Absolutely essential\"). Some topics have been abbreviated for clarity; the full topic names presented to the respondents can be found in the Supplementary Materials."
#| fig-height: 12

plot_likert(
  survey_long,
  all_topic_questions,
  counts = FALSE,
  labels = TRUE
) +
  theme(
    legend.position = "bottom",
    legend.direction = "vertical"
  )
```

Barriers to participate in SIB courses were also assessed using a 5-point Likert scale. The overall distribution of responses is shown in @fig-likert-barriers, which indicates that the majority of respondents did not perceive any of the listed factors as a significant barrier to their participation in SIB courses. However, some respondents indicated that distance and course fees could be potential barriers.

```{r}
#| label: fig-likert-barriers
#| fig-cap: "Overall responses of the question \"Please indicate how much each of the following factors would be a barrier for your participation in a SIB training course.\". The barriers are ordered by the proportion of respondents who indicated that they were not a barrier."
#| fig-height: 6
plot_likert(
  survey_long,
  "Please indicate how much each of the following factors would be a barrier for your participation in a SIB training course.",
  counts = FALSE,
  labels = TRUE
) +
  ggtitle(NULL) +
  theme(
    legend.position = "bottom",
    legend.direction = "vertical"
  )
```

When asked about their preferred learning formats, respondents expressed a preference for blended modes, multi-day intensive courses and interactive and problem-based learning approaches, such as Bring Your Own Data (BYOD) and Problem-Based Learning (PBL). Remarkably, AI-assisted formats were generally less favored. The overall distribution of responses is shown in @fig-likert-formats.

```{r}
#| label: fig-likert-formats
#| fig-cap: "Overall responses of the question \"Specify which learning formats you prefer. Note: these are not mutually exclusive, i.e. some can be used in combination\". The formats are ordered by the proportion of respondents who indicated that they liked or strongly liked the format."
#| fig-height: 6
plot_likert(
  survey_long,
  "Specify which learning formats you prefer Note: these are not mutually exclusive, i.e. some can be used in combination",
  counts = FALSE,
  labels = TRUE
) +
  ggtitle(NULL) +
  theme(
    legend.position = "bottom",
    legend.direction = "vertical"
  )
```

### Graph-based clustering of course preference responses

To group respondents based on their training needs, we constructed a network where nodes represent respondents and edges represent the similarity of their training preferences. Initially, we identified five clusters (@fig-course-cluster-network), of which one cluster (Cluster 5) consisted of a single respondent who indicated 'absolutely essential' for all course topics, which was considered as an outlier and therefore removed from the analysis.  

```{r}
#| label: cluster-course-preferences
#| fig-width: 4

course_scale <- meta_df |>
  dplyr::filter(main_question %in% all_topic_questions) |>
  dplyr::pull(scale) |>
  purrr::pluck(1)

course_matrix_tbl <- survey_long |>
  dplyr::filter(
    question_type == "likert",
    main_question %in% all_topic_questions,
    !is.na(sub_question),
    !is.na(answer)
  ) |>
  dplyr::mutate(
    feature = paste(main_question, sub_question, sep = " :: "),
    score = as.numeric(match(answer, course_scale))
  ) |>
  dplyr::select(respondent_id, feature, score) |>
  tidyr::pivot_wider(names_from = feature, values_from = score)

course_matrix <- course_matrix_tbl |>
  dplyr::select(-respondent_id) |>
  as.matrix()

# setting NAs to 1 (the lowest score, corresponding to "Not needed") allows us to keep respondents with some missing values in the correlation step, under the assumption that non-responses indicate a lack of interest in the topic.
course_matrix[is.na(course_matrix)] <- 1

# Pairwise respondent correlation network. NAs are handled pairwise.
respondent_cor <- stats::cor(t(course_matrix), use = "pairwise.complete.obs")
diag(respondent_cor) <- 0

# Keep only stronger positive edges to avoid a fully connected graph.
edge_threshold <- stats::quantile(
  respondent_cor[upper.tri(respondent_cor)],
  probs = 0.8,
  na.rm = TRUE
)

# create graph from adjacency matrix
adjacency <- respondent_cor
adjacency[is.na(adjacency)] <- 0
adjacency[adjacency < edge_threshold] <- 0

g <- igraph::graph_from_adjacency_matrix(
  adjacency,
  mode = "undirected",
  weighted = TRUE,
  diag = FALSE
)

# add respondent_id as vertex attribute for later merging with survey data
igraph::V(g)$respondent_id <- course_matrix_tbl$respondent_id

# louvain community detection with edge weights and resolution parameter to control cluster granularity.
set.seed(20260511)
community <- igraph::cluster_louvain(
  g,
  weights = igraph::E(g)$weight,
  resolution = 1
)

# create a tibble mapping respondent_id to cluster membership
cluster_assignments <- tibble::tibble(
  respondent_id = igraph::V(g)$respondent_id,
  cluster = factor(paste("Cluster", igraph::membership(community)))
)
```

```{r}
#| label: fig-course-cluster-network
#| fig-cap: "Respondent network from course-topic preference correlations. Edge widths indicate stronger positive correlation; colors indicate Louvain communities."
#| fig-height: 5

layout_fr <- igraph::layout_with_fr(g)

cluster_levels <- cluster_assignments |>
  dplyr::distinct(cluster) |>
  dplyr::arrange(cluster) |>
  dplyr::pull(cluster)

cluster_palette <- scales::hue_pal()(length(cluster_levels))
names(cluster_palette) <- cluster_levels

vertex_clusters <- paste("Cluster", igraph::membership(community))
vertex_colors <- unname(cluster_palette[vertex_clusters])

edge_weights <- igraph::E(g)$weight
edge_width <- scales::rescale(edge_weights, to = c(0.4, 2.5))

plot(
  g,
  layout = layout_fr,
  vertex.size = 4,
  vertex.label = NA,
  vertex.color = vertex_colors,
  edge.width = edge_width,
  edge.color = grDevices::adjustcolor("grey35", alpha.f = 0.35),
  margin = 0.05
)

legend(
  "topleft",
  legend = names(cluster_palette),
  col = unname(cluster_palette),
  pch = 16,
  pt.cex = 1.2,
  bty = "n",
  cex = 0.85
)
```

The structural topology of the resulting clusters displays varying levels of internal coherence. To evaluate the similarity of responses within the clusters, we calculated the edge density (the proportion of actual connections out of all possible links)for each cluster (@tbl-cluster-interconnectedness), showing  that Cluster 1 exhibits the highest edge density. Therefore, Cluster 1 has the most similar training needs.

```{r}
#| label: tbl-cluster-interconnectedness
#| tbl-cap: "Within-cluster interconnectedness scores (internal edge count and density)."

membership_vec <- igraph::membership(community)

cluster_interconnectedness <- sort(unique(membership_vec)) |>
  purrr::map_dfr(function(cl_id) {
    vids <- which(membership_vec == cl_id)
    subg <- igraph::induced_subgraph(g, vids = vids)

    # number of nodes and edges within the cluster
    n_nodes <- igraph::vcount(subg)
    # number of edges with both endpoints in the cluster
    internal_edges <- igraph::ecount(subg)
    # number of possible edges within the cluster (undirected, no self-loops)
    possible_edges <- n_nodes * (n_nodes - 1) / 2
    # total weight of edges within the cluster
    total_internal_weight <- if (internal_edges > 0) {
      sum(igraph::E(subg)$weight)
    } else {
      0
    }

    tibble::tibble(
      cluster = paste("Cluster", cl_id),
      nodes = n_nodes,
      `internal edges` = internal_edges,
      `possible edges` = possible_edges,
      # proportion of possible edges that actually exist within the cluster
      `edge density` = ifelse(
        possible_edges > 0,
        internal_edges / possible_edges,
        NA_real_
      ) #,
      # average number of internal connections per node within the cluster
      #   `mean internal degree` = mean(igraph::degree(subg)),
      #   `total internal weight` = total_internal_weight,
      #   `mean internal weight` = ifelse(internal_edges > 0, total_internal_weight / internal_edges, NA_real_)
    )
  })

cluster_interconnectedness |>
  dplyr::filter(cluster != "Cluster 5") |>
  dplyr::arrange(cluster) |>
  dplyr::mutate(
    `edge density` = round(`edge density`, 3) #,
    # `mean internal weight` = round(`mean internal weight`, 3)
  ) |>
  knitr::kable()
```

### Learner profiles

The four resulting clusters were characterized by distinct training needs, shown by their top 10 of most liked topics (@fig-topics-by-cluster). This clearly shows that responses within each cluster have distinct preferences that are not visible when analyzing the entire respondent pool (@fig-likert-preferences). Cluster 1 (@fig-topics-by-cluster A; referred to as 'AI and data') is characterized by an interest in AI and data topics, including application LLMs and reproducible and FAIR Data. Cluster 2 (@fig-topics-by-cluster B; referred to as 'Infectious disease genomics') is characterized by infectious disease topics and possibly ecology, with an interest in phylogenetics, pathogen genomics and epidemiology. Cluster 3 (@fig-topics-by-cluster C; referred to as 'Functional genomics') is characterized by an interest in functional genomics topics, with an interest in transcriptomics and epigenetics. Cluster 4 (@fig-topics-by-cluster D; referred to as 'Biochemistry, imaging & AI') is characterized by training needs on AI, biochemistry and imaging, with an interest in proteomics, mass spectrometry and imaging and omics data integration. All clusters have topics related to AI in their top 10, indicating that AI is a cross-cutting theme across all clusters.

```{r}
#| label: cluster-filtering

# get cluster sizes and filter out clusters with ≤ 5 respondents to ensure meaningful profiles in the next step
cluster_sizes_full <- cluster_assignments |>
  dplyr::count(cluster, name = "n_respondents") |>
  dplyr::arrange(cluster)

valid_clusters <- cluster_sizes_full |>
  dplyr::filter(n_respondents > 5) |>
  dplyr::pull(cluster)

all_clusters <- cluster_sizes_full |>
  dplyr::pull(cluster)

cluster_assignments_filtered <- cluster_assignments |>
  dplyr::filter(cluster %in% valid_clusters)

cluster_lookup <- cluster_assignments_filtered |>
  dplyr::mutate(cluster = as.character(cluster))

survey_results <- survey_results |>
  dplyr::mutate(respondent_id = dplyr::row_number()) |>
  dplyr::left_join(cluster_lookup, by = "respondent_id") |>
  dplyr::select(-respondent_id)

survey_long_cluster <- survey_long |>
  dplyr::inner_join(cluster_lookup, by = "respondent_id")
```

```{r}
#| label: fig-topics-by-cluster
#| fig-cap: "The top 10 most liked topics are shown for each cluster, indicating that each cluster has distinct training needs. Only the answers \"Very needed\" (light green) and \"Absolutely essential\" (dark green) are shown. The numbers in the bars represent the count of respondents within the respective cluster selecting the answer. (A) Cluster 1: AI and data. (B) Cluster 2: Infectious disease genomics. (C) Cluster 3: Functional genomics. (D) Cluster 4: Biochemistry, imaging & AI. Some topics have been abbreviated for clarity; the full topic names presented to the respondents can be found in the Supplementary Materials."
#| fig-height: 20

plot_list <- list()
for (cl in valid_clusters) {
  # cl_title <- if (cl %in% names(cluster_title_map)) cluster_title_map[[cl]] else cl

  p <- plot_likert(
    survey_long_cluster |> dplyr::filter(cluster == cl),
    all_topic_questions,
    counts = TRUE,
    labels = TRUE,
    positive_only = TRUE,
    top_n_sub_questions = 10
  ) 

  plot_list[[cl]] <- p
}

wrap_plots(plot_list, ncol = 1) +
  plot_annotation(tag_levels = "A") &
  theme(legend.position = "none")
```

These clusters can be further characerized based on demographic characteristics (@fig-cluster-pie-career; Supplementary Figure X), structural barriers to participation (Supplementary Figure X), and preferred learning formats (Supplementary Figure X). Such analyses can help training providers to design targeted curricula that address the specific needs of each cluster. For example, Cluster 1 (AI & data) has a higher proportion of early-career scientists (PhD candidates and postdoctoral researchers; @fig-cluster-pie-career) and a higher proportion of respondents from academia (Supplmentary Figure X), while Cluster 4 (Biochemistry, imaging & AI) has a higher proportion of senior scientists and respondents from industry. These differences in demographics can inform the design of training programs that are tailored to the specific needs and preferences of each cluster.

```{r}
#| label: cluster-pie-helpers

cluster_title_map <- c(
  "Cluster 1" = "Cluster 1: AI and Data",
  "Cluster 2" = "Cluster 2: Infectious Disease Genomics",
  "Cluster 3" = "Cluster 3: Functional Genomics",
  "Cluster 4" = "Cluster 4: Biochemistry, Imaging & AI"
)

cluster_display_labels <- vapply(
  valid_clusters,
  function(cl) {
    if (cl %in% names(cluster_title_map)) cluster_title_map[[cl]] else cl
  },
  character(1)
)

career_choices <- get_question_choices("Which career stage are you currently in?")
org_choices <- get_question_choices("How would you describe the organisation in which you currently work?")
country_choices <- get_question_choices("In which country do you work?")
canton_choices <- get_question_choices("In which canton do you work?")
sib_course_choices <- get_question_choices("When did you last attend a SIB course?")

cluster_demo <- survey_results |>
  dplyr::filter(cluster %in% valid_clusters)

cluster_demo <- cluster_demo |>
  dplyr::mutate(
    swiss_other = dplyr::case_when(
      `In which country do you work?` %in% c("Switzerland", "Suisse") ~ "Switzerland",
      `In which country do you work?` %in% country_choices ~ "Other",
      TRUE ~ NA_character_
    )
  )
```

```{r}
#| label: fig-cluster-pie-career
#| fig-cap: "Career stage composition per cluster."
#| fig-height: 6

career_pie <- make_cluster_pie_data(
  cluster_demo,
  value_col = "Which career stage are you currently in?",
  allowed_choices = career_choices,
  top_n = 5
)

plot_cluster_pies(career_pie, "Career stage by cluster") + ggtitle(NULL)
```

## Discussion

### Overall survey outcomes

Although the survey was open to anyone, the majority of respondents were from Switzerland and academia. The likely cause was that the survey was primarily advertised through the SIB training mailing list and other academic channels and networks within Switzerland. However, academic researchers working in Switzerland are the main target group of SIB training, and the survey results provide valuable insights into the training needs and preferences of bioinformatics researchers. Here, we discuss the overall survey outcomes along the three dimensions of training needs, participation barriers and learning formats.

#### Training needs

From the complete pool of survey results, it becomes clear that there is a demand for training on using (generative) AI and LLMs. The course that was indicated as most needed was "AI-assisted coding and analysis", and "Using large language models (LLM) for biomedical knowledge extraction". This shows that the respondents are aware of the potential of AI and LLMs in bioinformatics and experience a gap in their current skill set on how to use these tools effectively. 

The survey also revealed that there is a demand for training on FAIR data management and reproducible research, with "Making bioinformatics workflows findable and citable" and "Sharing and reusing life science data" as 3rd and 4th most needed respectively. This shows awareness of a need for learning about open and reproducible research practices amongst most repsondents, which corresponds with earlier findings regarding perceptions on FAIR data practices [@kersloot_perceptions_2022], and its challenges [@martinez-garcia_fair_2023]. 

While applying such outcomes to actual training, it is important to note that there can be a gap between intention - indicating a desire or need for training - and action - actually participating in a training course on the topic. Despite the questions on training needs were phrased in a way that the answer should relate to the respondents' own training needs (".. specify your training needs .."), some answers might represent what the respondents think their communities or colleagues need. This is also reflected in the relatively high proportion of senior scientists and PIs in the survey (25.7%), who might be more likely to answer on behalf of their group or lab rather than their own personal training needs. Therefore, the translation of these answers from intention to action, i.e. actually participating in a training course on the topic, might not be straightforward. 

#### Participation barriers

For most barriers, we saw that the majority of respondents did not perceive them as a barrier or as small barrier to their participation in SIB courses. Distance and course fees were large or critical barriers for a substantial proportion of respondents. 

Currently, approximately 50% of the SIB courses is provided online, and most other courses in-person in Lausanne, Geneva, Basel, Bern and Zurich. For online courses, the distance barrier should not be significant, also taking into account that barriers related to time-zone differences are minimal. Especially courses on emerging topics are held in-person, and usually in the city where the trainer is located. To releave the distance barrier for Swiss participants, there could be more incentive to organise the same course in different cities and possibly provide more online options. 

At the time of writing, participating in a regular SIB course costs CHF 100 (= $ 124) per day, which can be a substantial amount for some researchers, especially if they cannot receive funding from their organisation. Compared to other live courses with a similar target group in Europe (e.g. EMBL-EBI courses, VIB training), this is a moderate fee, and only covers a part of the actual costs. Rearranging the cost model could lower this barrier, but would also mean that either different choices should be made regarding the quality or quantity of courses, or that additional funding is required, e.g. through institutional memberships.

#### Learning formats

Respondents particularly liked short interactive live course formats, and combinations of both live and self-paced learning (blended). This shows the need of the human aspect in courses, such as the interaction between trainers and learners, and between learners themselves. 

There was a tendency towards dislike of AI-assisted learning formats, which is interesting given the high demand for training on AI and LLMs. This indicates that while respondents are interested in learning about AI, they prefer to learn in a more traditional format, with human interaction and guidance. How respondents interpreted AI-assisted learning formats is not clear, and could be interpreted in different ways. For example, it could be interpreted as a course where the trainer uses AI to assist in teaching, or as a course where the learners use AI to assist in learning. Implementation of AI-assisted learning formats could be done in different ways, and it is important to understand how learners would appreciate this format, and what their expectations are. Therefore, using AI in courses for the sake of just using AI is not recommended, and it is important to use it only for specific use-cases where it adds clear value.

### Learner profiles

The problems that can be solved with bioinformatics are very diverse, and professionals solving these problems are
 highly heterogeneous. Therefore, analyzing the survey as a homogeneous group, or slicing the group by demographic indicators only partly captures the diversity of training needs. This issue is not specific to bioinformatics training, creating clusters of respondents is an overall valuable approach for understanding survey data, for example in citizenship norms [@hooghe_comparative_2016], health literacy [@cheng_measuring_2024; @hyde_electronic_2019], bullying [@chung_are_2020] or company training [@wenger_analysing_2024]. 
 
 By using graph-based clustering of the answers on training topics, we got a much more granular view, resulting in insights that would remain hidden otherwise. For example, the topic 'AI-assisted coding and analysis' was indicated as most needed by the complete pool of respondents, but when looking at the clusters, it was only in the most needed topic in Cluster 1 (AI & Data; 1st most needed), and not even in the top 10 for cluster 3 (Functional genomics). This shows that organising a course on this topic would serve the primary needs of Cluster 1, but only partly of respondents in the other clusters. This works the other way around as well. For example, the topic "Phylogenetics data analysis" was the 15th most needed topic in the complete pool of respondents, indicating a moderately to low priority to organize a course on the topic. However, it was the first most needed topic in Cluster 2 (Infectious disease genomics). This shows that organising a course on this topic would serve the primary needs of a substantial group of respondents, and this would not be visible when looking at the complete pool of respondents.

The clusters also enable a more targeted approach to address barriers to participation, learning formats and demographics. For example, Cluster 1 (AI & Data) has a higher proportion of early-career scientists  and a higher proportion of respondents from Swiss universities, while Cluster 4 (Biochemistry, imaging & AI) has a higher proportion of senior scientists and respondents from industry. These differences in demographics can inform the design of training programs that are tailored to the specific needs and preferences of each cluster.

## Conclusion

This study demonstrates the utility of network modeling and community detection for transforming raw training needs survey data into clear, actionable curriculum design. By grouping respondents based on their shared technical requirements rather than classical demographic labels, we isolated four stable, cohesive learner archetypes. The heavy influence of AI across these clusters—both as a highly requested technical topic and as an overall disliked pedagogical format—underscores the need for agile curriculum design. Tailoring training curricula to these specific learner profiles allows computational biology training providers to effectively address participation barriers, adopt optimal teaching formats, and ensure maximum educational impact in a rapidly evolving scientific landscape.
